---
author: "Emil Westin"
format: gfm
execute: 
  eval: false
editor: 
  markdown: 
    wrap: sentence
---


## Updating R

The following code assumes you are using windows and can be run from RStudio.
The version of R can usually be found here `C:\Program Files\R\R-4.2.0`.
[More info](https://cran.r-project.org/bin/windows/base/rw-FAQ.html#What_0027s-the-best-way-to-upgrade_003f "R for Windows FAQ").

```{r}
install.packages("installr")
library(installr)
# Run the following 
# (you can also just type installr::updateR() and choose options interactively)
installr::updateR(browse_news = T, 
                  fast = F,
                  install_R = T, 
                  copy_packages = T, 
                  copy_site_files = T,
                  keep_old_packages = T,
                  update_packages = F, # this will be done in next step
                  start_new_R = F,
                  quit_R = F
                  )
# When the above is completed, restart R, exit R studio and reopen with a fresh session

# Reinstall the copied packages to match the latest version of R
update.packages(checkBuilt=TRUE, ask=FALSE) 

# Double check that the packages have been built with the latest version of R:
installed.packages()
```


## Unload library/package from R session without restarting

```{r}
# unloading library(officer)
detach(package:officer)
```


## RStudio useful shortcuts

|                                                                            | Command                            |
|----------------------------------------------------------------------------|------------------------------------|
| Restart R-session and unload packages:                                     | `Command`/`Ctrl` + `Shift` + `F10` |
| Comment/uncomment single/multiple lines:                                   | `Ctrl`+ `Shift` + `c`              |
| Display command palette:                                                   | `Ctrl`+ `Shift` + `p`              |
| Auto-indent:                                                               | `Ctrl` + `i`                       |
| Edit multiple lines simultaneusly (increase the blinking line vertically): | `Ctrl` + `Alt` + `up / down`       |
| Duplicate current line:                                                    | `Alt` + `Shift` + `up / down`      |
| Insert assignment (`<-`):                                                  | `Alt` + `-`                        |
| Insert pipe operator (`%>%`):                                              | `Ctrl` + `Shift` + `m`             |
| Delete current line:                                                       | `Ctrl` + `d`                       |
| Multiple word selection:                                                   | `Ctrl` + `Alt` + double left click |
| Run code from beginning to line                                            | `Ctrl` + `Alt` + `b`               |
| Move/shift line up/down                                                    | `Alt` + `up / down`                |
